---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

## chandwich: Chandler-Bate Sandwich Loglikelihood Adjustment

### What does chandwich do?

The `chandwich` package performs adjustments of an independence loglikelihood using a robust sandwich estimator of the parameter covariance matrix, based on the methodology in \href{http://dx.doi.org/10.1093/biomet/asm015}{Chandler and Beta (2007)}. This can be used for cluster correlated data when interest lies in the parameters of the marginal distributions.

### A simple example

The main function in the threshr package is `adjust_loglik`.  It finds the maximum likelihood estimate (MLE) of model parameters based on an independence loglikelihood in which cluster dependence in the data is ignored.  The independence loglikelihood is adjusted in a way that ensures that the Hessian of the adjusted loglikelihood conicides with a robuest sandwich estimate of the parameter covariance at the MLE.  Three adjustments are available: one in which the independence loglikelihood itself is scaled (vertical scaling) and two others where the scaling is in the parameter vector (horizontal scaling).

```{r, eval = FALSE}
```

### Installation

To get the current released version from CRAN:

```{r installation, eval = FALSE}
install.packages("chandwich")
```

### Vignette

See `vignette("threshr-vignette", package = "threshr")` for an overview of the package.

